\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-10-15 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
compiled paper with updated cs1

```{r erspc-cox, eval = eval_cs1, echo=FALSE}
surv_obj <- with(ERSPC, Surv(Follow.Up.Time, DeadOfPrCa))
cox_model <- survival::coxph(surv_obj ~ ScrArm, data = ERSPC)
```

```{r erspc-cox-cif, eval = FALSE, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk functions for control and screening groups in the ERSPC data. We plot the estimate from the Cox model using \\code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line). We used a cubic B-spline expansion of time."}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
myCols <- cbPalette[c(4,7)]

@@

```{r}
knitr::knit_exit()
```

>>>>

```{r erspc-cox, eval = eval_cs1, echo=FALSE}
surv_obj <- with(ERSPC, Surv(Follow.Up.Time, DeadOfPrCa))
cox_model <- survival::coxph(surv_obj ~ ScrArm, data = ERSPC)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
myCols <- cbPalette[c(4,7)]
```

```{r erspc-cox-cif, eval = FALSE, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk functions for control and screening groups in the ERSPC data. We plot the estimate from the Cox model using \\code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line). We used a cubic B-spline expansion of time."}

@@

### 2021-10-15 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
compiled paper with updated cs1

For our first case study, we return to the ERSPC study and investigate
the differences in risk between the control and screening arms. We fit a
flexible hazard by using a smooth function of time modeled with cubic
B-splines. The model is fit using \code{fitSmoothHazard} with the
familiar formula interface:

\begin{Schunk}
\begin{Sinput}
library(splines)
fit <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
                       data = ERSPC, ratio = 100)
\end{Sinput}
\end{Schunk}

The output object from \code{fitSmoothHazard} inherits from the
\code{singleEventCB} and \code{glm} classes. For this reason, we can
leverage the \code{summary} method for \code{glm} objects to output a
familiar summary of the results:

\begin{Schunk}
\begin{Sinput}
summary(fit) 
\end{Sinput}
\begin{Soutput}
#> Fitting smooth hazards with case-base sampling
#> 
#> Sample size: 159893 
#> Number of events: 540 
#> Number of base moments: 54000 
#> ----
#> 
#> Call:
#> fitSmoothHazard(formula = DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
#>     data = ERSPC, ratio = 100)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.403  -0.172  -0.135  -0.085   3.801  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value Pr(>|z|)    
#> (Intercept)           -10.2533     0.3168  -32.36  < 2e-16 ***
#> bs(Follow.Up.Time)1     4.2544     0.7976    5.33  9.6e-08 ***
#> bs(Follow.Up.Time)2     2.1371     0.4720    4.53  5.9e-06 ***
#> bs(Follow.Up.Time)3     4.5339     0.6889    6.58  4.7e-11 ***
#> ScrArmScreening group  -0.2481     0.0886   -2.80   0.0051 ** 
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059.0  on 54539  degrees of freedom
#> Residual deviance: 5787.6  on 54535  degrees of freedom
#> AIC: 5798
#> 
#> Number of Fisher Scoring iterations: 8
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- c(0, 1)

absoluteRisk(fit, time = new_time, newdata = new_data)
\end{Sinput}
\begin{Soutput}
#>  time                
#>     0 0.0e+00 0.0e+00
#>     1 5.4e-05 4.2e-05
\end{Soutput}
\end{Schunk}

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated risk
function from \pkg{casebase} on the Cox model estimate. As noted in the
Theoretical Details section, the usual asymptotic results hold for
likelihood ratio tests built using case-base sampling models. Therefore,
we can easily test the significance of the spline model:

\begin{Schunk}
\begin{Soutput}
#> Analysis of Deviance Table
#> 
#> Model: binomial, link: logit
#> 
#> Response: DeadOfPrCa
#> 
#> Terms added sequentially (first to last)
#> 
#> 
#>                    Df Deviance Resid. Df Resid. Dev Pr(>Chi)    
#> NULL                               54539       6059             
#> bs(Follow.Up.Time)  3      263     54536       5796   <2e-16 ***
#> ScrArm              1        8     54535       5788   0.0048 ** 
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{Soutput}
\end{Schunk}

The null hypothesis is rejected in favor of the spline model. Similarly,
to compare different models (e.g.~time modeled linearly), we could
compute Akaike's Information Criterion (AIC) for each model and compare
them.

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cox-cif-1} \caption{Risk functions for control and screening groups in the ERSPC data. We plot the estimate from the Cox model using \code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line). We used a cubic B-spline expansion of time.}\label{fig:erspc-cox-cif}
\end{figure}
\end{Schunk}

Next, we can compute confidence intervals for the risk function using
the method \code{confint.absRiskCB}.

\begin{Schunk}
\begin{Sinput}
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
conf_ints <- confint(risk, fit)
conf_ints
\end{Sinput}
\begin{Soutput}
#>   time estimate conf.low conf.high        cov_prof
#> 1    0  0.0e+00  0.0e+00   0.0e+00   Control group
#> 2    1  5.4e-05  3.5e-05   9.3e-05   Control group
#> 3    0  0.0e+00  0.0e+00   0.0e+00 Screening group
#> 4    1  4.2e-05  2.7e-05   7.3e-05 Screening group
\end{Soutput}
\end{Schunk}

These point-wise confidence intervals are constructed using parametric
bootstrap; the validity of this approach relies on the asymptotic
normality of the MLE. In Figure \ref{fig:erspc-cif-conf}, we see the
95\% confidence bands around the estimates.

@@

Previous re-analyses of these data suggest that the 20\% reduction in
prostate cancer death due to screening was an underestimate
\citep{hanley2010mortality}. The estimated 20\% (from a proportional
hazards model) did not account for the delay between screening and the
time the effect is expected to be observed. As a result, the null
effects in years 1--7 masked the substantial reductions that began to
appear from year 8 onward. This motivates the use of a time-dependent
hazard ratio which can easily be fit with the \pkg{casebase} package by
including an interaction term with time in the model:

\begin{Schunk}
\begin{Sinput}
fit_inter <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) * ScrArm, 
                             data = ERSPC)
\end{Sinput}
\end{Schunk}

Note that we did not have to specify any cut points, as would be the
case with the \code{survSplit} function in the \pkg{survival} package.
In Figure \ref{fig:interaction-ERSPC}, we have the estimated hazard
ratio and 95\% confidence interval for screening vs.~control group as a
function of time using the \code{plot} method for objects of class
\code{singleEventCB}:

\begin{Schunk}
\begin{Sinput}
plot(fit_inter, type = "hr", newdata = new_data,
     var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
\end{Sinput}
\end{Schunk}

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./interaction-ERSPC-1} \caption[Estimated hazard ratio and 95\% confidence interval for screening vs]{Estimated hazard ratio and 95\% confidence interval for screening vs. control group as a function of time in the ERSPC dataset. Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic B-spline basis of follow-up time and treatment arm. 95\% confidence intervals are calculated using the delta method. The plot shows that the effect of screening only begins to become statistically apparent by year 7. The 25-60\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\% reported in the original article.}\label{fig:interaction-ERSPC}
\end{figure}
\end{Schunk}

@@

\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
\centering
\begin{tabular}[t]{lrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Covariates & HR & 95\% Conf. & HR & 95\% Conf.\\
\midrule
Sex & 0.74 & (0.43, 1.3) & 0.68 & (0.39, 1.21)\\
Disease & 0.52 & (0.29, 0.94) & 0.52 & (0.28, 0.94)\\
Phase (CR2 vs. CR1) & 1.20 & (0.48, 3) & 1.21 & (0.47, 3.09)\\
Phase (CR3 vs. CR1) & 1.59 & (0.41, 6.17) & 1.67 & (0.46, 6.08)\\
Phase (Relapse vs. CR1) & 4.09 & (1.9, 8.81) & 4.55 & (1.98, 10.46)\\
\addlinespace
Source & 1.70 & (0.55, 5.23) & 1.46 & (0.47, 4.53)\\
Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
\bottomrule
\end{tabular}
\end{table}

>>>>

For our first case study, we return to the ERSPC study and investigate
the differences in risk between the control and screening arms. Previous
re-analyses of these data suggest that the 20\% reduction in prostate
cancer death due to screening was an underestimate
\citep{hanley2010mortality}. The estimated 20\% (from a proportional
hazards model) did not account for the delay between screening and the
time the effect is expected to be observed. As a result, the null
effects in years 1--7 masked the substantial reductions that began to
appear from year 8 onward. This motivates the use of a time-dependent
hazard ratio which can easily be fit with the \pkg{casebase} package by
including an interaction term with time in the model. We fit a flexible
hazard by using a smooth function of time modeled with a penalized cubic
spline basis with 2 degrees of freedom (implemented in the
\code{survival::pspline} function). The model is fit using
\code{fitSmoothHazard} with the familiar formula interface:

\begin{Schunk}
\begin{Sinput}
fit <- fitSmoothHazard(DeadOfPrCa ~ pspline(Follow.Up.Time, df = 2) * ScrArm, 
                       data = ERSPC, ratio = 100)
\end{Sinput}
\end{Schunk}

Note that we did not have to specify any cut points, as would be the
case with the \code{survSplit} function in the \pkg{survival} package.
The output object from \code{fitSmoothHazard} inherits from the
\code{singleEventCB} and \code{glm} classes. For this reason, we can
leverage the \code{summary} method for \code{glm} objects to output a
familiar summary of the results:

\begin{Schunk}
\begin{Sinput}
summary(fit) 
\end{Sinput}
\begin{Soutput}
#> Fitting smooth hazards with case-base sampling
#> 
#> Sample size: 159893 
#> Number of events: 540 
#> Number of base moments: 54000 
#> ----
#> 
#> Call:
#> fitSmoothHazard(formula = DeadOfPrCa ~ pspline(Follow.Up.Time, 
#>     df = 2) * ScrArm, data = ERSPC, ratio = 100)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.477  -0.164  -0.140  -0.084   3.882  
#> 
#> Coefficients:
#>                                                        Estimate Std. Error
#> (Intercept)                                             -12.613      9.756
#> pspline(Follow.Up.Time, df = 2)1                          1.320     10.705
#> pspline(Follow.Up.Time, df = 2)2                          5.190      9.526
#> pspline(Follow.Up.Time, df = 2)3                          4.481      9.866
#> pspline(Follow.Up.Time, df = 2)4                          6.060      9.690
#> pspline(Follow.Up.Time, df = 2)5                          5.231      9.851
#> pspline(Follow.Up.Time, df = 2)6                          9.176      9.749
#> pspline(Follow.Up.Time, df = 2)7                         -0.434     19.572
#> ScrArmScreening group                                     7.900     13.050
#> pspline(Follow.Up.Time, df = 2)1:ScrArmScreening group   -7.962     14.507
#> pspline(Follow.Up.Time, df = 2)2:ScrArmScreening group   -8.351     12.690
#> pspline(Follow.Up.Time, df = 2)3:ScrArmScreening group   -7.787     13.224
#> pspline(Follow.Up.Time, df = 2)4:ScrArmScreening group   -8.098     12.942
#> pspline(Follow.Up.Time, df = 2)5:ScrArmScreening group   -9.389     13.223
#> pspline(Follow.Up.Time, df = 2)6:ScrArmScreening group   -7.385     13.114
#> pspline(Follow.Up.Time, df = 2)7:ScrArmScreening group  -17.380     30.806
#>                                                        z value Pr(>|z|)
#> (Intercept)                                              -1.29     0.20
#> pspline(Follow.Up.Time, df = 2)1                          0.12     0.90
#> pspline(Follow.Up.Time, df = 2)2                          0.54     0.59
#> pspline(Follow.Up.Time, df = 2)3                          0.45     0.65
#> pspline(Follow.Up.Time, df = 2)4                          0.63     0.53
#> pspline(Follow.Up.Time, df = 2)5                          0.53     0.60
#> pspline(Follow.Up.Time, df = 2)6                          0.94     0.35
#> pspline(Follow.Up.Time, df = 2)7                         -0.02     0.98
#> ScrArmScreening group                                     0.61     0.54
#> pspline(Follow.Up.Time, df = 2)1:ScrArmScreening group   -0.55     0.58
#> pspline(Follow.Up.Time, df = 2)2:ScrArmScreening group   -0.66     0.51
#> pspline(Follow.Up.Time, df = 2)3:ScrArmScreening group   -0.59     0.56
#> pspline(Follow.Up.Time, df = 2)4:ScrArmScreening group   -0.63     0.53
#> pspline(Follow.Up.Time, df = 2)5:ScrArmScreening group   -0.71     0.48
#> pspline(Follow.Up.Time, df = 2)6:ScrArmScreening group   -0.56     0.57
#> pspline(Follow.Up.Time, df = 2)7:ScrArmScreening group   -0.56     0.57
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059.0  on 54539  degrees of freedom
#> Residual deviance: 5772.7  on 54524  degrees of freedom
#> AIC: 5805
#> 
#> Number of Fisher Scoring iterations: 9
\end{Soutput}
\end{Schunk}

As noted in the Theoretical Details section, the usual asymptotic
results hold for likelihood ratio tests built using case-base sampling
models. Therefore, we can easily test the significance of the spline
term and its interaction with time:

\begin{Schunk}
\begin{Sinput}
anova(fit, test = "LRT")
\end{Sinput}
\begin{Soutput}
#> Analysis of Deviance Table
#> 
#> Model: binomial, link: logit
#> 
#> Response: DeadOfPrCa
#> 
#> Terms added sequentially (first to last)
#> 
#> 
#>                                        Df Deviance Resid. Df Resid. Dev
#> NULL                                                   54539       6059
#> pspline(Follow.Up.Time, df = 2)         7    269.1     54532       5790
#> ScrArm                                  1      8.0     54531       5782
#> pspline(Follow.Up.Time, df = 2):ScrArm  7      9.1     54524       5773
#>                                        Pr(>Chi)    
#> NULL                                               
#> pspline(Follow.Up.Time, df = 2)          <2e-16 ***
#> ScrArm                                   0.0046 ** 
#> pspline(Follow.Up.Time, df = 2):ScrArm   0.2454    
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{Soutput}
\end{Schunk}

Similarly, to compare different models (e.g.~time modeled linearly), we
could compute Akaike's Information Criterion (AIC) for each model and
compare them.

\hypertarget{time-dependent-hazard-ratios}{%
\subsubsection{Time-dependent hazard
ratios}\label{time-dependent-hazard-ratios}}

In what follows, the hazard ratio for a variable \(X\) is defined as

\[
\frac{\lambda\left(t | X=x_1, \mathbf{Z}=\mathbf{z_1} ; \hat{\beta}\right)}{\lambda(t | X=x_0, \mathbf{Z}=\mathbf{z_0} ; \hat{\beta})}
\] where \(\lambda(t|\cdot;\hat{\beta})\) is the hazard rate as a
function of the variable \(t\) (which is usually time, but can be any
other continuous variable), \(x_1\) is the value of \(X\) for the
exposed group, \(x_0\) is the value of \(X\) for the unexposed group,
\(\mathbf{Z}\) are other covariates in the model which are equal to
\(\mathbf{z_1}\) in the exposed and \(\mathbf{z_0}\) in the unexposed
group, and \(\hat{\beta}\) are the estimated regression coefficients. As
indicated by the formula above, it is most instructive to plot the
hazard ratio as a function of a variable \(t\) only if there is an
interaction between \(t\) and \(X\). Otherwise, the resulting plot will
simply be a horizontal line across time.

The \code{plot} method with \code{type="hr"} for objects of class
\code{singleEventCB} can be used to compute time-dependent hazard ratios
and confidence intervals. In Figure \ref{fig:interaction-ERSPC}, we show
the estimated hazard ratio and 95\% confidence interval for screening
vs.~control group as a function of time. Note that we must specify the
covariate profiles and times for the predicted hazards.

\begin{Schunk}
\begin{Sinput}
new_time <- seq(1, 12, by  = 0.1)
new_data <- data.frame(ScrArm = factor("Control group",
                                         levels = c("Control group","Screening group")),
                      Follow.Up.Time = new_time)
plot(fit, type = "hr", newdata = new_data,
     var = "ScrArm", xvar = "Follow.Up.Time", ci = T)
\end{Sinput}
\end{Schunk}

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./interaction-ERSPC-1} \caption[Estimated hazard ratio and 95\% confidence interval for screening vs]{Estimated hazard ratio and 95\% confidence interval for screening vs. control group as a function of time in the ERSPC dataset. Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic pspline basis of follow-up time and treatment arm. 95\% confidence intervals are calculated using the delta method. The plot shows that the effect of screening only begins to become statistically apparent by year 7. The 25-60\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\% reported in the original article.}\label{fig:interaction-ERSPC}
\end{figure}
\end{Schunk}

\hypertarget{hazard-functions}{%
\subsubsection{Hazard functions}\label{hazard-functions}}

Modeling the hazard function directly allows us to easily visualize it
with the \code{plot} method and \code{type="hazard"} for objects of
class \code{singleEventCB}. We plot the hazard functions for both
treatment arms in Figure \ref{fig:cs1hazard}. The pattern we see is
consistent with the population-time plot shown in Figure
\ref{fig:plot-erspc-data} C, where more events are observed at later
follow-up times. The drop at the end can be explained by the fact that
very few observations were followed for the entire 15 year period.

\begin{Schunk}
\begin{Sinput}
plot(fit, type = "hazard",
     hazard.params = list(xvar = "Follow.Up.Time",
                          by = "ScrArm"))
\end{Sinput}
\begin{Soutput}
#> Conditions used in construction of plot
#> ScrArm: Control group / Screening group
#> offset: 0
\end{Soutput}

\includegraphics[width=\textwidth,keepaspectratio=true]{./cs1hazard-1} \end{Schunk}

\hypertarget{absolute-risk}{%
\subsubsection{Absolute risk}\label{absolute-risk}}

@@

\begin{Schunk}
\begin{Sinput}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14,0.1)
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
head(risk)
\end{Sinput}
\begin{Soutput}
#>  time                
#>   0.0 0.0e+00 0.0e+00
#>   0.1 2.0e-06 6.2e-06
#>   0.2 4.2e-06 1.2e-05
#>   0.3 6.6e-06 1.8e-05
#>   0.4 9.2e-06 2.3e-05
#>   0.5 1.2e-05 2.9e-05
\end{Soutput}
\end{Schunk}

We can subsequently compute confidence intervals for the risk function
using the method \code{confint.absRiskCB}:

\begin{Schunk}
\begin{Sinput}
conf_ints <- confint(risk, fit)
head(conf_ints)
\end{Sinput}
\begin{Soutput}
#>   time estimate conf.low conf.high      cov_prof
#> 1  0.0  0.0e+00  0.0e+00   0.0e+00 Control group
#> 2  0.1  2.0e-06  2.5e-07   1.6e-05 Control group
#> 3  0.2  4.2e-06  6.1e-07   3.0e-05 Control group
#> 4  0.3  6.6e-06  1.1e-06   4.2e-05 Control group
#> 5  0.4  9.2e-06  1.8e-06   5.2e-05 Control group
#> 6  0.5  1.2e-05  2.8e-06   6.2e-05 Control group
\end{Soutput}
\end{Schunk}

These point-wise confidence intervals are constructed using parametric
bootstrap; the validity of this approach relies on the asymptotic
normality of the MLE. In Figure \ref{fig:erspc-cif-conf}, we see the
95\% confidence bands around the estimates. We also overlay the
Kaplan-Meier curves as a reference.

@@

@@

\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
\centering
\begin{tabular}[t]{lrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Covariates & HR & 95\% Conf. & HR & 95\% Conf.\\
\midrule
Sex & 0.75 & (0.43, 1.31) & 0.68 & (0.39, 1.21)\\
Disease & 0.55 & (0.3, 0.99) & 0.52 & (0.28, 0.94)\\
Phase (CR2 vs. CR1) & 1.10 & (0.44, 2.74) & 1.21 & (0.47, 3.09)\\
Phase (CR3 vs. CR1) & 1.75 & (0.45, 6.78) & 1.67 & (0.46, 6.08)\\
Phase (Relapse vs. CR1) & 4.23 & (1.96, 9.14) & 4.55 & (1.98, 10.46)\\
\addlinespace
Source & 1.61 & (0.53, 4.91) & 1.46 & (0.47, 4.53)\\
Age & 1.00 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
\bottomrule
\end{tabular}
\end{table}
